---
title: "Ugalde_linear_reg"
author: "Axela Keeza T. Ugalde"
date: "`r Sys.Date()`"
output: html_document
---
# Activity 11 - Linear Regression {.tabset .tabset-fade .tabset-pills}
  
## Data Preparation
```{r setup, echo = TRUE, warning = TRUE, fig.align = "center"}
library(sf)
library(terra)
library(dplyr)
library(ggplot2)
library(ggnewscale)
library(tidyverse)

# Read CSV
my_data <- read.csv("herpetozoa-033-095-s001.csv")

# Clean the data by removing NA
clean_data <- my_data[complete.cases(my_data),]

# Countercheck if any "NA" is found
sum(is.na(clean_data))

# Elevation
phil_dem <- rast("ph_ strm_geo.tif")

# Subset coordinates
# Clean and Subset
coord_data <- clean_data |>
  select(Long, Lat)

points_vect <- vect(coord_data, 
                      geom = c("Long","Lat"), 
                      crs = crs(phil_dem))

# Extract elev of coords
elev_data <- terra::extract(phil_dem, points_vect)
elev_data

# Combine and Rename
regression_data <- data.frame(richness = clean_data$SPPTotal,
                              elevation = elev_data$Band_1)
regression_data
```

## A. Simple linear regression
```{r manual, echo = TRUE, warning = TRUE, fig.align = "center"}

# Data
x_indep <- regression_data$elevation
y_dep <- regression_data$richness

# Means
xbar <- mean(x_indep)
ybar <- mean(y_dep)

# Slope (b)
b <- sum((x_indep - xbar)*(y_dep - ybar))/sum((x_indep - xbar)^2)
b

# Intercept (a)
a <- ybar - b*xbar
a

# Write the equation
cat("Regression Equation:\n")
cat("y =", round(a,3), "+", round(b,3), "x\n")

# Calculate the predicted arm length
yhat <- a + b * x_indep

data.frame(
  Elevation = x_indep,
  SPPTotal = y_dep,
  Predicted = round(yhat,2)
)
## Plot the data with fitted line
plot(x_indep, y_dep,
     pch = 19,
     col = "palevioletred1",
     xlab = "Elevation (m)",
     ylab = "Species Richness/SPPTotal",
     main = "Simple Linear Regression")

abline(a, b,
       col = "maroon",
       lwd = 2)

## Plot with predicted values
pdf("simplelinearReg_elev_richness.pdf", width = 10, height = 8)

## Observed data
my_plot <- plot(x_indep, y_dep,
     pch = 19,
     col = "palevioletred1",
     xlab = "Elevation (m)",
     ylab = "Species Richness/SPPTotal",
     main = "Simple Linear Regression")

## Add fitted regression line
abline(a, b,
       col = "maroon",
       lwd = 2)

## Add predicted values
points(x_indep, yhat,
       pch = 17,      # Triangle
       col = "orchid2",
       cex = 1.2)

## Draw residual lines
segments(x0 = x_indep, y0 = y_dep,
         x1 = x_indep, y1 = yhat,
         lty = 2,
         col = "gray50")

## Add legend
legend("topright",
       legend = c("Observed", "Predicted", "Regression Line"),
       pch = c(19, 17, NA),
       lty = c(NA, NA, 1),
       col = c("palevioletred1", "orchid2", "maroon"),
       lwd = c(NA, NA, 2),
       bty = "n")
dev.off()

## Calculate R squared
## Sum of Squared Errors (SSE)
## SSE = Σ(y - ŷ)²

SSE <- sum((y_dep - yhat)^2)

## Total Sum of Squares (SST)
## SST = Σ(y - ȳ)²

SST <- sum((y_dep - ybar)^2)

## Coefficient of Determination
## R² = 1 - (SSE / SST)

R2 <- 1 - (SSE / SST)
R2
```
There is a **negative linear relationship** between elevation and species richness.

## B. Multiple linear regression
```{r mult, echo = TRUE, warning = TRUE, fig.align = "center"}

# Load and Extract
precip_rast <- rast("annual_precipitation.tif")
temp_rast <- rast("annual_temperature.tif")

precip_data <- terra::extract(precip_rast, points_vect)
temp_data <- terra::extract(temp_rast, points_vect)

# Combine data
combined_data <- regression_data |>
  mutate(precipitation = precip_data[ , 2],
         temperature = temp_data[ ,2])

# Multiple regression formula: y ~ x1 + x2 + x3
multiple_lreg <- lm(richness ~ elevation + precipitation + temperature, data = combined_data)

# Display the full results
summary(multiple_lreg)
```

*Which relationship is statistically significant?*

A relationship is considered significant if this value is less than 0.05.
Based from the summary table, the **precipitation** with a p-value of 0.02770 and **temperature** with p-value of 0.03107.
