---
title: "My First Rmarkdown - Activity 3 (Summary Stats)"
author: "Axela Keeza T. Ugalde"
date: "01 June 2026"
output: html_document
---
# Activity 3 - Summary statistics, frequency distribution, and basic graphical representation


## A. Read and perform descriptive statistics
### A.1 Data Loading and Cleaning
```{r readcsv, echo = TRUE, warning = TRUE, fig.align = "center"}
# Read the csv file
my_data <- read.csv("herpetozoa-033-095-s001.csv")

# Check class of data
class(my_data)

# Clean the data by removing NA
clean_data <- my_data[complete.cases(my_data),]

# Countercheck if any "NA" is found
sum(is.na(clean_data))
```
### A.2 Show Data Structure
```{r datastr, echo = TRUE, warning = TRUE, fig.align = "center"}
# Show the data structure
str(clean_data)
```
### A.3 Elevation Summary Statistics
Subset the elevation data to calculate the mean, standard deviation, and its range.
```{r elev, echo = TRUE, warning = TRUE, fig.align = "center"}
# Subset data
elev_data <- clean_data$Elev

library(pander)
elev_mean <- mean(elev_data)
elev_sd <- sd(elev_data)
elev_range <- range(elev_data)
elev_descstat <- cbind(elev_mean, elev_sd, elev_range[1], elev_range[2])
elev_descstat
colnames(elev_descstat) <- c("Mean", "Standard Deviation", "Range Min", "Range Max")
pander(elev_descstat)

```
### A.4 Species Observations
Using 2 loop methods to count observations for species in Columns G to AG from the csv file.

#### Loop Methods {.tabset .tabset-fade .tabset-pills}

##### Typical Loop
```{r typLoop, echo = TRUE, warning = TRUE, fig.align = "center"}
## Subset data
species_obs <- clean_data[ , 7:33]
species_name <- colnames(species_obs)

#List to store the data per iteration of the Typical Loop
species_tot <- vector("list", length = 27)
for (i in 1:27){
species_sum <- sum(species_obs[ , i])
species_tot[[i]] <- species_sum
names(species_tot)[i] <- species_name[i]

species_tot
}
```

##### Compact Loop by Sapply
```{r CompLoop, echo = TRUE, warning = TRUE, fig.align = "center"}
species_totCL <- sapply(species_obs, sum)
species_totCL
```

#### A.4 Which species has the most and the fewest observations?
```{r qs_A4, echo = TRUE, warning = TRUE, fig.align = "center"}
most_obsSpe <- max(species_totCL)
species_totCL[species_totCL == most_obsSpe]
fewest_obsSpe <- min(species_totCL)
species_totCL[species_totCL == fewest_obsSpe]
```

### A.5 Habitat Types
```{r htypes, echo = TRUE, warning = TRUE, fig.align = "center"}
h_type <- cbind(clean_data["Htypes"], clean_data["Phlo"], clean_data["Babu"], clean_data["Geat"], clean_data["Hety"], clean_data["Bosc"], clean_data["Drru"], clean_data["Apbo"])

# Split for easier calculations
htype_split <- split(h_type, h_type$Htypes)
htype_split
```
### A.5 Determine the mean count of these species in each habitat type.
```{r htypes_mean, echo = TRUE, warning = TRUE, fig.align = "center"}
htype_mean <- sapply(htype_split, function(x){
colMeans(x[ , 2:8])
})
pander(htype_mean)
```

### A.5 Which habitat type has the most species counts?
```{r htypes_sum, echo = TRUE, warning = TRUE, fig.align = "center"}
hytpe_sum <- sapply(htype_split, function(x){
colSums(x[ ,2:8])
})
pander(hytpe_sum)
htype_tot <- sapply(htype_split, function(x){
sum(x[ ,2:8])
})
pander(htype_tot)
maxHtype <- max(htype_tot)
htype_tot[htype_tot == maxHtype]
```
## B. Plots

### B.1 Scatterplot
```{r scatterplot, echo = TRUE, warning = TRUE, fig.align = "center"}
plot_data <- cbind(clean_data["SPPTotal"], clean_data["Elev"], clean_data["Htypes"])
plot_data

plot_xy <- plot(clean_data$Elev, clean_data$SPPTotal,
xlab = "Elevation",
ylab = "Species Total Count",
main = "Herpetofauna Distribution by Elevation",
col = as.factor(plot_data$Htypes),
pch = 16
)
legend("topright",
legend = levels(as.factor(plot_data$Htypes)),
col = 1:length(levels(as.factor(plot_data$Htypes))),
pch = 16)
```

#### B1. Are there habitat types with no species observations? What are these?
```{r qs_b1, echo = TRUE, warning = TRUE, fig.align = "center"}

plot_htype_tot <- tapply(plot_data$SPPTotal, plot_data$Htypes, sum)
plot_htype_tot
#None
```
### B.2 Bar Plot
```{r barplot, echo = TRUE, warning = TRUE, fig.align = "center"}
barplot_xy <- barplot(plot_htype_tot,
main = "Total Species Observed per Habitat",
xlab = "Habitat Types",
ylab = "Total Species Observed",
col = 1:4)
```


### B.3 Histogram
```{r histogram, echo = TRUE, warning = TRUE, fig.align = "center"}

hist_xy <- hist(plot_data$SPPTotal,
main = "Histogram of Total Species Counts",
xlab = "Total Species Counts",
ylab = "Frequency",
col = "lightpink",
border = "black",
breaks = 8)
```

## C. Interpretation

 Based on the scatterplot, for species total count v elevation, there was a clear observation of a negative relationship between the elevation and species richness. Elevations below 500 have higher biodiversity(6-8). Meanwhile, higher elevations of above 500 have 2 or less. Meanwhile, for species total count vs habitat type, most of the OGF are with higher elevation but was observed with 1-5 species, ASG and ESG sites were on the lower elevation but has higher observed species. Meanwhile, CVT habitat type had low species total count even with the varying elevation.
Barplots of total species observed per habitat showed that the habitat type "ASG" has the highest cumulative species observed (58), followed by "OGF" (48), then "ESG"(28), and lastly "CVT"(1).
The histogram of total species count shows a positively skewed distribution indicating more observation sites are species-poor and there were less sites with high-diversity.
