---
title: "My First Rmarkdown - Activity 4 (Tidyverse)"
author: "Axela Keeza T. Ugalde"
date: "01 June 2026"
output: html_document
---
  
# Activity 4 - Tidverse: Data wrangling
  
## A. Read and Filter Out {.tabset .tabset-fade .tabset-pills}
### A.1 Data Loading
```{r dataload, echo = TRUE, warning = TRUE, fig.align = "center"}
library(pander)
library(tidyverse)

# Read data
my_data <- read.csv("habitat_types.csv")
my_data

# Check type of class
class(my_data)

# Check Data Structure and NA
sum(is.na(my_data))
my_data |> glimpse()
```

### A.2 Subset Data
```{r subset, echo = TRUE, warning = TRUE, fig.align = "center"}
# Subset data - Transect, Htypes, Small Trees, Flowering Trees
select_data <- select(my_data, Transect, Htypes, SmallTrees, FloweringTrees)
pander(select_data)

## Filter CVT and Sort "Small Trees" (descending)
select_data |>
  filter(Htypes == "CVT") |>
  arrange(desc(SmallTrees)) ->
  cvt_descSTrees

pander(cvt_descSTrees)
```

### A.3 Transect with Most Small Trees
```{r qs_a2, echo = TRUE, warning = TRUE, fig.align = "center"}
cvt_descSTrees |> 
  filter(SmallTrees == max(SmallTrees)) -> mostStrees

pander(mostStrees)
```

### A.4 Mean of Flowering Trees/Htypes
```{r mean, echo = TRUE, warning = TRUE, fig.align = "center"}
ftrees_htype <- select_data |> 
  group_by(Htypes) |> 
  summarize(mean_flowering = mean(FloweringTrees))

pander(ftrees_htype)
```

**Which habitat types do not differ significantly in the mean number of flowering trees?**

ASG and ESG do not differ significantly in the mean number of flowering trees with 5.26 and 5.38, respectively. Meanwhile, CVT has 8.86.

### A.5 HerbGroundDiff - Difference bet HerbsCover and GroundCover using Mutate
```{r hgrounddiff, echo = TRUE, warning = TRUE, fig.align = "center"}
sel_df_mod <- select_data |> 
  mutate(HerbGroundDiff = my_data$HerbsCover - my_data$GroundCover) |>
  arrange(HerbGroundDiff)

pander(sel_df_mod)
```

### A.6 Transect with No Difference
```{r zeroes, echo = TRUE, warning = TRUE, fig.align = "center"}
sel_df_mod |> 
  filter(HerbGroundDiff == 0) -> zeroes

pander(zeroes)
```

### A.7 Wide Version
```{r widever, echo = TRUE, warning = TRUE, fig.align = "center"}
sel_df_modWide <- sel_df_mod |>
                  pivot_wider(
                    names_from = Htypes,
                    values_from = c(SmallTrees, FloweringTrees, HerbGroundDiff),
                  )
view(sel_df_modWide)
pander(sel_df_modWide)
```


# B. Reflection


Compared to the last activity, this activity was faster to understand. It made subsetting and applying functions/operation with filtering and sorting easier.
I just need more time to practice the transformations but the pipe function makes things more efficient and clean-looking.