---
title: "Ugalde_PCA"
author: "Axela Keeza T. Ugalde"
date: "`r Sys.Date()`"
output: html_document
---

# Activity 9 - Dimension Reduction: Principal Component Analysis (PCA) {.tabset .tabset-fade .tabset-pills}

## A. Principal Component Analysis {.tabset .tabset-fade .tabset-pills}
### Data Preparation
```{r setup, echo = TRUE, warning = TRUE, fig.align = "center"}

# Data Preparation
library(tidyverse)
library(PCAtest)

# Data
my_data <- read.csv("Meneses_et.al_2026_Platymantis.csv")
as.data.frame(my_data)

# Subsetting using pipeline
subset_dat <- my_data |> 
  select(Species , 10:ncol(my_data)) |>
  drop_na()

subset_dat
sum(is.na(subset_dat))
  
# Check class and structure
class(subset_dat)
glimpse(subset_dat)

## How many morphological features are present in the dataset?
25
```


### Cleaning, Scaling, Correlation
```{r CCS, echo = TRUE, warning = TRUE, fig.align = "center"}

# Clean
clean_dat <- subset_dat[complete.cases(subset_dat), ]

# Scaling

numeric_dat <- clean_dat |>
  select(where(is.numeric))
scaled_dat <- scale(numeric_dat)

scaled_dat

# Correlations
cor_dat <- cor(scaled_dat)
cor_dat
```


### Eigenvectors and %Proportions
```{r calc, echo = TRUE, warning = TRUE, fig.align = "center"}

# Eigenvectors and eigenvalues
eig <- eigen(cor_dat)
eig

# Proportion of variance in percentage
prop_var <- eig$values / sum(eig$values)
prop_var

# Percentage
per_prop <- prop_var*100
per_prop

# Vectors
eig_vec <- eig$vectors
scores <- scaled_dat %*% eig_vec

head(scores)

## How many principal components were produced? 
num_pcs <- length(eig$values)
num_pcs

## What is the proportion of variance explained by components 1–4?
sum_4pcs <- sum(per_prop[1:4])
sum_4pcs
```

### Screeplot and PCA Biplot
```{r plots, echo = TRUE, warning = TRUE, fig.align = "center", fig.width = 18, fig.height = 8}
par(mfrow = c(1, 2))

barplot(prop_var,
        names.arg = paste0("PC", 1:length(prop_var)),
        ylim = c(0, 1),
        col = "lightpink",
        main = "Explained Variance (Scree Plot)",
        ylab = "Proportion of Variance")

clean_dat$Species <- as.factor(clean_dat$Species)

## Get eigenvectors for PC1 and 2
loadings <- eig$vectors[, 1:2]

## Scale to match PC scores
load_scale <- 3  # adjust for visibility
loadings_scaled <- loadings * load_scale

## Recreate biplot
plot(scores[,1], scores[,2],
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA Score Plot with Eigenvector Loadings",
     pch = 19,
     col = clean_dat$Species)

## Add points legend
legend("topright",
       legend = levels(clean_dat$Species),
       col = 1:3,
       pch = 19)

## Add eigenvector arrows
arrows(0, 0,
       loadings_scaled[,1],
       loadings_scaled[,2],
       col = "black",
       lwd = 2)

## Add labels for variables
text(loadings_scaled[,1],
     loadings_scaled[,2],
     labels = colnames(numeric_dat),
     col = "black",
     pos = 3)
```

## B. Broken-stick vs PCAtest {.tabset .tabset-fade .tabset-pills}

What is the difference between these two methods for assessing the importance of principal components?

The two methods differ by the level of randomization utilizied to create a null model (or the randomly generated data points) for checking is the empirical data (observed data) captured a true biological pattern or a random noise.

Broken-Stick Test breaks the eigenvectors into smaller pieces and randomly samples these to calculate expected value for each component. On the other hand, the PCAtest permutates the raw dataset (or observed data), runs full PCA for every single randomization, and evaluates the scores and proportion of variance against the empirical data.

### Examine the variance - Broken-Stick Test
```{r brokenstick, echo = TRUE, warning = TRUE, fig.align = "center"}
# Broken-stick Function
p <- ncol(numeric_dat)

## Function
broken_stick <- function(p){
  sapply(1:p, function(k){
    sum(1 / (k:p)) / p
  })
}

bs <- broken_stick(ncol(my_data[,1:length(prop_var)]))
bs

## Combine observed and broken stick variance
comparison <- data.frame(
  PC = paste0("PC", 1:length(prop_var)),
  Observed = prop_var,
  BrokenStick = bs
)
comparison

## Plot
plot(prop_var,
     type = "b",
     pch = 19,
     ylim = c(0, max(prop_var)),
     xlab = "Principal Component",
     ylab = "Proportion of Variance",
     main = "Broken-Stick Analysis")

lines(bs,
      type = "b",
      pch = 17,
      col = "red")

legend("topright",
       legend = c("Observed", "Broken Stick"),
       pch = c(19, 17),
       col = c("black", "red"))
```

### Examine the variance - PCAtest
```{r PCAtest, echo = TRUE, warning = FALSE, fig.align = "center"}
result <- PCAtest(scaled_dat, 100, 100, 0.05, 
                  varcorr=FALSE, 
                  counter=FALSE, 
                  plot=TRUE)
```